# Snippets from revised chapters

## Likelihood-Ratio Test {#Chap-03-06-model-comparison-LR-test}

The likelihood-ratio (LR) test is a very popular frequentist method of model comparison. The LR-test assimilates model comparison to frequentist hypothesis testing. It defines a suitable test statistic and supplies an approximation of the sampling distribution. The LR-test first and foremost applies to the comparison of **nested models**, but there are results about how approximate results can be obtained when comparing non-nested models with an LR-test [@Vuong1989:Likelihood-Rati]. 

A frequentist model $M_i$ is **nested** inside another frequentist model $M_j$ iff $M_i$ can be obtained from $M_j$ by fixing at least one of $M_j$'s free parameters to a specific value. If $M_i$ is nested under $M_j$, $M_i$ is called the nested model, and $M_j$ is called the **nesting model** or the **encompassing model**. Obviously, the nested model is simpler (of lower complexity) than the nesting model.

For example, we had the two-parameter exponential model previously:

$$
\begin{aligned}
P(D = \langle k, N \rangle \mid \langle a, b\rangle) & = \text{Binom}(k,N, a \exp (-bt)), \ \ \ \ \text{where } a,b>0 
\end{aligned}
$$

An example of a model that is nested under this two-parameter model is the following one-parameter model, which fixes $a = 1.1$.

$$
\begin{aligned}
P(D = \langle k, N \rangle \mid b) & = \text{Binom}(k,N, 1.1 \ \exp (-bt)), \ \ \ \ \text{where } b>0 
\end{aligned}
$$

Here's an ML-estimation for the nested model:

```{r}
nLL_expo_nested <- function(b) {
  # calculate predicted recall rates for given parameters
  theta <- 1.1*exp(-b*t)  # one-param exponential model 
  # avoid edge cases of infinite log-likelihood
  theta[theta <= 0.0] <- 1.0e-4
  theta[theta >= 1.0] <- 1-1.0e-4
  # return negative log-likelihood of data
  - sum(dbinom(x = obs, prob = theta, size = 100, log = T))
}

bestExpo_nested <- optim(
  nLL_expo_nested, 
  par = 0.5, 
  method = "Brent", 
  lower = 0, 
  upper = 20
)
bestExpo_nested
```

The LR-test looks at the likelihood ratio of the nested model $M_0$ over the encompassing model $M_1$ using the following test statistic:

$$\text{LR}(M_1, M_0) = -2\log \left(\frac{P_{M_0}(D_\text{obs} \mid \hat{\theta}_0)}{P_{M_1}(D_\text{obs} \mid \hat{\theta}_1)}\right)$$

We can calculate the value of this test statistic for the current example as follows:

```{r}
LR_observed <- 2 * bestExpo_nested$value - 2 * bestExpo$value
LR_observed
```

If the simpler (nested) model is true, the sampling distribution of this test statistic approximates a $\chi^2$-distribution with $d$ if we have more and more data. The degrees of freedom $d$ is given by the difference in free parameters, i.e., the number of parameters the nested model fixes to specific values, but which are free in the nesting model. 

We can therefore calculate the $p$-value for the LR-test for our current example like so:

```{r}
p_value_LR_test <- 1 - pchisq(LR_observed, 1)
p_value_LR_test
```

The $p$-value of this test quantifies the evidence against the assumption that the data was generated by the simpler model. A significant test result would therefore indicate that it would be surprising if the data was generated by the simpler model. This is standardly taken as evidence in favor of the more complex, nesting model. For the current $p$-value  $p \approx `r p_value_LR_test %>% round(4)`$, there is no strong evidence against the simpler model and we would therefore rather favor the nested model due to its simplicity; the data at hand does not seem to warrant the added complexity of the nesting model; the nested model seems to suffice.



<div class = "exercises">
**Exercise 11.2**

TRUE OR FALSE?

a. The nested model usually has more free parameters than the nesting model.
b. When we perform the LR-test, we initially assume that the nested model is more plausible.
c. A LR-test can only compare the nested model with nesting models.
d. If the LR-test result has a $p$-value equal to 1.0, one can conclude that it's a piece of evidence in favor of the simpler model. 

<div class="collapsibleSolution">
<button class="trigger">Solution</button>
<div class="content">

a. False
b. True
c. False
d. True

</div>
</div>
</div>

