# Snippets from revised chapter "Parameter estimation"

## Table comparing frequentist and Bayesian estimates

Parameter estimation is traditionally governed by two measures: (i) a point-estimate for the best parameter value, and (ii) an interval-estimate for a range of values that are considered "good enough". Table \@ref(tab:ch-03-03-estimation-overview) gives the most salient answers that each approach gives.

```{r ch-03-03-estimation-overview, echo = F}
table_data <- tribble(
  ~estimate, ~Bayesian, ~frequentist,
  "best value", "mean of posterior  posterior", "maximum likelihood estimate",
  "interval range", "credible interval (HDI)", "confidence interval"
)
knitr::kable(
  table_data,
  escape = F,
  caption = "Common methods of obtaining point-valued and interval-range estimates for parameters, given some data, in frequentist and Bayesian approaches.", 
  booktabs = TRUE
)
```

## Comparing Bayesian and frequentist estimates {#ch-03-03-estimation-comparison}

For Bayesians, point-valued and interval-based estimates are just summary statistics to efficiently communicate about or reason with the main thing: the full posterior distribution. For the frequentist, the point-valued and interval-based estimates might be all there is. Computing a full posterior can be very hard. Computing point-estimates is usually much simpler. Yet, all the trouble of having to specify priors, and having to calculate a much more complex mathematical object, can pay off. An example which is intuitive enough is that of a likelihood function in a multi-dimensional parameter space where there is an infinite collection of parameter values that maximize the likelihood function (think of plateau). Asking a godly oracle for the (actually: an) MLE can be disastrously misleading. The full posterior will show the quirkiness.^[We will see such an example later for the case of linear regression with correlated independent variables, so-called collinearity.]

Practical issues aside, there are also conceptual arguments that can be pinned against each other. Suppose you do not know the bias of a coin, you flip it once and it lands heads. The case in mathematical notation: $k=1$, $N=1$. As a frequentist, your "best" estimate of the coin's bias is that it is 100% rigged: it will *never* land tails. As a Bayesian, with uninformed priors, your "best" estimate is, following Laplace rule, $\frac{k+1}{N+2} = \frac{2}{3}$. Notice that there might be different notions of what counts as "best" in place. Still, the frequentist "best" estimate seems rather extreme.

What about interval-ranged estimates? Which is the better tool, confidence intervals or credible intervals? -- This is harder to answer. Numerical simulations can help answer these questions.^[Even if the math seems daunting, this method is much more tangible and applicable and requires only basic programming experience.] The idea is simple but immensely powerful. We simulate, repeatedly, a ground-truth and synthetic results for fictitious experiments, and then we apply the statistical tests/procedures to these fictitious data sets. Since we know the ground-truth, we can check which tests/procedures got it right.

Let's look at a simulation, comparing credible intervals to confidence intervals, the latter of which are calculated by asymptotic approximation or the so-called exact method. To do so, we repeatedly sample a ground-truth (e.g., a known coin bias $\theta_{\text{true}}$) from a flat distribution over $[0;1]$.^[This is already not innocuous. We are fixing, as it were, an assumption about how likely ground-truths should actually occur in the real world.]. We then simulate an experiment in a synthetic world with $\theta_{\text{true}}$, using a fixed value of $n$, here taken from the set $n \in \set{10, 25, 100, 1000}$. We then construct a confidence interval (either approximately or precisely) and a 95% credible interval; for each of the three interval estimates. We check whether the ground-truth $\theta_{\text{true}}$ is *not* included in any given interval estimate. We calculate the mean number of times such as non-inclusion (errors!) happen for each kind of interval estimate. The code below implements this and the figure below shows the results based on 10,000 samples of $\theta_{\text{true}}$.

```{r}
# how many "true" thetas to sample
n_samples <- 10000 
# sample a "true" theta
theta_true <- runif(n=n_samples)
# create data frame to store results in
results <- expand.grid(
  theta_true = theta_true,
  n_flips = c(10,25,100, 1000)
) %>% 
  as_tibble() %>% 
  mutate(
    outcome = 0,
    norm_approx = 0,
    exact = 0,
    Bayes_HDI = 0
  )
  
for (i in 1:nrow(results)) {
  
  # sample fictitious experimental outcome for current true theta
  results$outcome[i] <- rbinom(
    n = 1, 
    size = results$n_flips[i], 
    prob = results$theta_true[i]
  )
  
  # get CI based on asymptotic Gaussian
  norm_approx_CI <- binom::binom.confint(
    results$outcome[i], 
    results$n_flips[i], 
    method = "asymptotic"
  )
  results$norm_approx[i] <- !(
    norm_approx_CI$lower <= results$theta_true[i] && 
      norm_approx_CI$upper >= results$theta_true[i]
    )
  
  # get CI based on exact method
  exact_CI <- binom::binom.confint(
    results$outcome[i], 
    results$n_flips[i], 
    method = "exact"
  )
  results$exact[i] <- !(
    exact_CI$lower <= results$theta_true[i] && 
      exact_CI$upper >= results$theta_true[i]
  )
  
  # get 95% HDI (flat priors)
  Bayes_HDI <- binom::binom.bayes(
    results$outcome[i], 
    results$n_flips[i], 
    type = "highest", 
    prior.shape1 = 1, 
    prior.shape2 = 1
  )
  results$Bayes_HDI[i] <- !(
    Bayes_HDI$lower <= results$theta_true[i] && 
      Bayes_HDI$upper >= results$theta_true[i]
  )
}

results %>% 
  gather(key = "method", "Type_1", norm_approx, exact, Bayes_HDI) %>% 
  group_by(method, n_flips) %>% 
  dplyr::summarize(avg_type_1 = mean(Type_1)) %>% 
  ungroup() %>% 
  mutate(
    method = factor(
      method, 
      ordered = T, 
      levels = c("norm_approx", "Bayes_HDI", "exact")
    )
  ) %>% 
  ggplot(aes(x = as.factor(n_flips), y = avg_type_1, color = method)) + 
  geom_point() + geom_line(aes(group = method)) +
  xlab("number of flips per experiment") +
  ylab("proportion of exclusions of true theta")


```

<div class = "exercises">
**Exercise 9.5** 

Pick the correct answer:

The most frequently used point-estimate of Bayesian parameter estimation looks at...

a. ...the median of the posterior distribution.

b. ...the maximum likelihood estimate.

c. ...the mean of the posterior distribution.

d. ...the normalizing constant in Bayes rule.

<div class="collapsibleSolution">
<button class="trigger">Solution</button>
<div class="content">

Statement c. is correct.

</div>
</div>

The most frequently used interval-based estimate in frequentist approaches is...

a. ...the support of the likelihood distribution.

b. ...the confidence interval.

c. ...the hypothesis interval.

d. ...the 95% highest-density interval of the maximum likelihood estimate.

<div class="collapsibleSolution">
<button class="trigger">Solution</button>
<div class="content">

Statement b. is correct.

</div>
</div>
</div>


## Best fits for linear regression

```{r}
# function for the negative log-likelihood of the given
# data and fixed parameter values
nll = function(y, x, beta_0, beta_1, sd) {
  # negative sigma is logically impossible
  if (sd <= 0) {return( Inf )}
  # predicted values
  yPred = beta_0 + x * beta_1
  # negative log-likelihood of each data point 
  nll = -dnorm(y, mean=yPred, sd=sd, log = T)
  # sum over all observations
  sum(nll)
}
fit_lh = optim(
  # initial parameter values
  par = c(1.5, 0, 0.5),
  # function to optimize
  fn = function(par) {
    with(avocado_data, 
         nll(average_price, total_volume_sold,
             par[1], par[2], par[3])
    )
  }
)
fit_lh$par
```

This result tells us that the best fitting parameter triple has an intercept of $\beta_0 \approx 1.42$, a slope $\beta_1 \approx -2.47$ and a standard deviation $\sigma \approx 0.39$. We can compare these values with a built-in function for linear regression models (which, however, does not return an estimate of $\sigma$ (see Chapter \@ref(Chap-04-01-simple-linear-regression) for more information)):

```{r}
lm(average_price ~ total_volume_sold, avocado_data)$coef
```
