--- 
title: "Introduction to Data Analysis"
author: ""
date: "last rendered at: `r Sys.time()`"
site: bookdown::bookdown_site
output:
  bookdown::tufte_html_book:
    tufte_variant: envisioned
  bookdown::gitbook:
    css: styles.css
    highlight: kate
    fig_caption: TRUE
    config:
      toc:
        collapse: section
      download: ["epub", "pdf"]
      sharing: no
  bookdown::tufte_book2:
    highlight: kate
    number_sections: yes
    citation_package: natbib
    dev: "cairo_pdf"
    includes:
      in_header: la.tex
  bookdown::epub_book: default
documentclass: scrbook
bibliography: [book.bib, packages.bib]
biblio-style: apalike
link-citations: yes
description: "Course Material for I2DA-2019 in Cognitive Science at University Osnabrück"
github-repo: "michael-franke/intro-data-analysis"
---

```{r init, echo=FALSE, message=FALSE, warning=FALSE, include=FALSE}
library(dplyr)
library(ggplot2)

options(tinytex.engine_args = '-shell-escape') # to appease the warning spirits
knitr::opts_chunk$set(warning = FALSE, message = FALSE)

# hrbrthemes::import_public_sans()
theme_set(hrbrthemes::theme_ipsum(plot_margin = margin(10, 10, 10, 10)))
```

# Preface {-}

The book introduces key concepts of data analysis from a frequentist and a Bayesian tradition. It uses R to handle, plot and analyze data. It relies on simulation to illustrate selected statistical concepts.


## Testing / Showcasing

Don't pay too much attention to what is written here.

### Quotes

This is a quote:

> Tidy datasets [...] have a specific structure: each variable is a column, each observation is a row, and each type of observational unit is a table.
>
> --- @wickham2014

### Infobox

At certain stages, possibly at the end of chapters or after important concepts, we might want to use a special infobox (see `.infobox` in `styles.css`) to summarise it or give food for thought. Like this:

```{block, type='infobox'}
A horse walks into a bar and orders a pint. The barkeep says "you're in here pretty often. Think you might be an alcoholic?", to which the horse says "I don't think I am.", and vanishes from existence.

See, the joke is about Descartes' famous philosophy of 'I think therefore, I am", but to explain that part before the rest of the joke would be to put Descartes before the horse.
```

We can have boxes with different icons for different purposes:

```{block, type='questions'}
This might be useful for excercises or general questions. Do you like it?
```

```{block, type='beware'}
Sometimes there are things that are really important, like exceptions to general rules. This box might be appropriate for these.
```

```{block2, type='research'}
["More research needs to be done"](https://en.wikipedia.org/wiki/Further_research_is_needed) as an infobox.
```

### Plots

This is a plot of quite a famous dataset [@anscombe1973]:

```{r plot_example_01}
tibble(
  grp = rep(c("I", "II", "III", "IV"), each = 11),
  x = c(anscombe$x1, anscombe$x2, anscombe$x3, anscombe$x4),
  y = c(anscombe$y1, anscombe$y2, anscombe$y3, anscombe$y4)
) %>% 
  ggplot(aes(x, y)) +
    geom_smooth(method = lm, se = F) +
    geom_point(color = "orange", size = 2) +
    scale_y_continuous(breaks = scales::pretty_breaks()) +
    scale_x_continuous(breaks = scales::pretty_breaks()) +
    labs(title = "Anscombe's Quartet", x = NULL, y = NULL,
         subtitle = bquote(y == 0.5 * x + 3 ~ (R^2 %~~% .667) ~ "for all datasets")) +
    facet_wrap(~grp, ncol = 2, scales = "free_x") +
    theme(strip.background = element_rect(fill = "#f2f2f2", colour = "white"))
```
