
# General description

In 2013, students of the Statistics class at FSEV UK were asked to invite their friends to participate in this survey.


**Year:** 
2013

**Citation of downloaded data set:** 
??

**Data collection mode:**
structured questionnaire

# Data set - description

The data file consists of 1010 rows and 150 columns (139 integer and 11 categorical). For convenience, the original variable names were shortened in the data file. See the [variable-description-yps.csv](../variable-description-yps.csv) file if you want to match the data with the original names.
    
The survey was presented to participants in both electronic and written form. The original questionnaire was in Slovak language and was later translated into English. All participants were of Slovakian nationality, aged between 15-30.

**Related files:**

- [overview-yps-dataset.Rmd](../overview-yps-dataset.Rmd) - overview of variable groups and variables
- [young-people-survey.csv](../young-people-survey.csv) - complete dataset
- [variable-description-yps.csv](../variable-description-yps.csv) - helpful overview and description of variable names (e.g. for filtering, etc.)

**Number of participants:** 
1010 respondents

**Missing data:**
Yes, but no special coding (labeld by NA)

## Overview variable groups

- Music preferences (19 items)
- Movie preferences (12 items)
- Hobbies & interests (32 items)
- Phobias (10 items)
- Health habits (3 items)
- Personality traits, views on life, & opinions (57 items)
- Spending habits (7 items)
- Demographics (10 items)
 
### load dataset and variable descriptions

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
# read data set
data <- read.csv("young-people-survey.csv")
# read variable descriptions
variables <- read.csv("variable-description-yps.csv")
# add third column for "group variable"
variables <- mutate(variables, Group = c(rep("group music",19),rep("group movies",12),rep("group interests",32), rep("group phobia",10),rep("group health",3),rep("group personality",57), rep("group spending",7), rep("group demographics",10)))
```

### Music preference

```{r}
filter(variables, Group=="group music") %>% 
  print()
```

#### Questionnaire:

- I enjoy listening to music.: Strongly disagree 1-2-3-4-5 Strongly agree (integer)
- I prefer.: Slow paced music 1-2-3-4-5 Fast paced music (integer)
- Dance, Disco, Funk: Don't enjoy at all 1-2-3-4-5 Enjoy very much (integer)
- Folk music: Don't enjoy at all 1-2-3-4-5 Enjoy very much (integer)
- Country: Don't enjoy at all 1-2-3-4-5 Enjoy very much (integer)
- Classical: Don't enjoy at all 1-2-3-4-5 Enjoy very much (integer)
- Musicals: Don't enjoy at all 1-2-3-4-5 Enjoy very much (integer)
- Pop: Don't enjoy at all 1-2-3-4-5 Enjoy very much (integer)
- Rock: Don't enjoy at all 1-2-3-4-5 Enjoy very much (integer)
- Metal, Hard rock: Don't enjoy at all 1-2-3-4-5 Enjoy very much (integer)
- Punk: Don't enjoy at all 1-2-3-4-5 Enjoy very much (integer)
- Hip hop, Rap: Don't enjoy at all 1-2-3-4-5 Enjoy very much (integer)
- Reggae, Ska: Don't enjoy at all 1-2-3-4-5 Enjoy very much (integer)
- Swing, Jazz: Don't enjoy at all 1-2-3-4-5 Enjoy very much (integer)
- Rock n Roll: Don't enjoy at all 1-2-3-4-5 Enjoy very much (integer)
- Alternative music: Don't enjoy at all 1-2-3-4-5 Enjoy very much (integer)
- Latin: Don't enjoy at all 1-2-3-4-5 Enjoy very much (integer)
- Techno, Trance: Don't enjoy at all 1-2-3-4-5 Enjoy very much (integer)
- Opera: Don't enjoy at all 1-2-3-4-5 Enjoy very much (integer)

### Movie preferences

```{r}
filter(variables, Group=="group movies") %>% 
  print()
```

#### Questionnaire:

-  I really enjoy watching movies.: Strongly disagree 1-2-3-4-5 Strongly agree (integer)
-  Horror movies: Don't enjoy at all 1-2-3-4-5 Enjoy very much (integer)
-  Thriller movies: Don't enjoy at all 1-2-3-4-5 Enjoy very much (integer)
-  Comedies: Don't enjoy at all 1-2-3-4-5 Enjoy very much (integer)
-  Romantic movies: Don't enjoy at all 1-2-3-4-5 Enjoy very much (integer)
-  Sci-fi movies: Don't enjoy at all 1-2-3-4-5 Enjoy very much (integer)
-  War movies: Don't enjoy at all 1-2-3-4-5 Enjoy very much (integer)
-  Tales: Don't enjoy at all 1-2-3-4-5 Enjoy very much (integer)
-  Cartoons: Don't enjoy at all 1-2-3-4-5 Enjoy very much (integer)
-  Documentaries: Don't enjoy at all 1-2-3-4-5 Enjoy very much (integer)
-  Western movies: Don't enjoy at all 1-2-3-4-5 Enjoy very much (integer)
-  Action movies: Don't enjoy at all 1-2-3-4-5 Enjoy very much (integer)

### Subjective well-being, social exclusion, religion, national and ethnic identity

```{r}
filter(variables, Group=="group interests") %>% 
  print()
```

#### Questionnaire:

-    History: Not interested 1-2-3-4-5 Very interested (integer)
-    Psychology: Not interested 1-2-3-4-5 Very interested (integer)
-    Politics: Not interested 1-2-3-4-5 Very interested (integer)
-    Mathematics: Not interested 1-2-3-4-5 Very interested (integer)
-    Physics: Not interested 1-2-3-4-5 Very interested (integer)
-    Internet: Not interested 1-2-3-4-5 Very interested (integer)
-    PC Software, Hardware: Not interested 1-2-3-4-5 Very interested (integer)
-    Economy, Management: Not interested 1-2-3-4-5 Very interested (integer)
-    Biology: Not interested 1-2-3-4-5 Very interested (integer)
-    Chemistry: Not interested 1-2-3-4-5 Very interested (integer)
-    Poetry reading: Not interested 1-2-3-4-5 Very interested (integer)
-    Geography: Not interested 1-2-3-4-5 Very interested (integer)
-    Foreign languages: Not interested 1-2-3-4-5 Very interested (integer)
-    Medicine: Not interested 1-2-3-4-5 Very interested (integer)
-    Law: Not interested 1-2-3-4-5 Very interested (integer)
-    Cars: Not interested 1-2-3-4-5 Very interested (integer)
-    Art: Not interested 1-2-3-4-5 Very interested (integer)
-    Religion: Not interested 1-2-3-4-5 Very interested (integer)
-    Outdoor activities: Not interested 1-2-3-4-5 Very interested (integer)
-    Dancing: Not interested 1-2-3-4-5 Very interested (integer)
-    Playing musical instruments: Not interested 1-2-3-4-5 Very interested (integer)
-    Poetry writing: Not interested 1-2-3-4-5 Very interested (integer)
-    Sport and leisure activities: Not interested 1-2-3-4-5 Very interested (integer)
-    Sport at competitive level: Not interested 1-2-3-4-5 Very interested (integer)
-    Gardening: Not interested 1-2-3-4-5 Very interested (integer)
-    Celebrity lifestyle: Not interested 1-2-3-4-5 Very interested (integer)
-    Shopping: Not interested 1-2-3-4-5 Very interested (integer)
-    Science and technology: Not interested 1-2-3-4-5 Very interested (integer)
-    Theatre: Not interested 1-2-3-4-5 Very interested (integer)
-    Socializing: Not interested 1-2-3-4-5 Very interested (integer)
-    Adrenaline sports: Not interested 1-2-3-4-5 Very interested (integer)
-    Pets: Not interested 1-2-3-4-5 Very interested (integer)

[WORK OUT LATER!]

### Climate change

```{r}
filter(variables, Group=="Group Climate change") %>% 
  print()
```

### Welfare attitudes

```{r}
filter(variables, Group=="Group Welfare attitudes") %>% 
  print()
```

### Gender, Year of birth and Household grid

```{r}
filter(variables, Group=="Group Gender, Year of birth and Household grid") %>% 
  print()
```

### Socio-demographics

```{r}
filter(variables, Group=="Group Socio-demographics") %>% 
  print()
```

### Human values

```{r}
filter(variables, Group=="Group Human values") %>% 
  print()
```