
# General description

The [World Values Survey](www.worldvaluessurvey.org) is a global network of social scientists studying changing values and their impact on social and political life. The survey, which started in 1981, seeks to use the most rigorous, high-quality research designs in each country. The WVS consists of nationally representative surveys conducted in almost 100 countries which contain almost 90 percent of the world's population, using a common questionnaire. The WVS is the largest non-commercial, cross-national, time series investigation of human beliefs and values ever executed, currently including interviews with almost 400,000 respondents. Moreover the WVS is the only academic study covering the full range of global variations, from very poor to very rich countries, in all of the world's major cultural zones.

**Actual wave:** 
The WVS is launching Wave 6 data covering the 2010-2014 wave. With 60 countries and societies around the world and more than 85,000 respondents, this is the latest ressource made available for the research community.

Wave 6 (2010-2014)

**Citation of downloaded data set:** 
Inglehart, R., C. Haerpfer, A. Moreno, C. Welzel, K. Kizilova, J. Diez-Medrano, M. Lagos, P. Norris, E. Ponarin & B. Puranen et al. (eds.). 2014. World Values Survey: Round Six - Country-Pooled Datafile Version: www.worldvaluessurvey.org/WVSDocumentationWV6.jsp. Madrid: JD Systems Institute. 

**Data collection mode:**
structured questionnaire

# Data set - description

The data file consists of 1010 rows and 150 columns (139 integer and 11 categorical). For convenience, the original variable names were shortened in the data file. See the [variable-description-yps.csv](../variable-description-yps.csv) file if you want to match the data with the original names.
    
The survey was presented to participants in both electronic and written form. The original questionnaire was in Slovak language and was later translated into English. All participants were of Slovakian nationality, aged between 15-30.

**Related files:**

- [overview-wvs-dataset.Rmd](../overview-wvs-dataset.Rmd) - overview of variable groups and variables
- [Official-questionnaire](../WV6_Official_Questionnaire.pdf) 
- [Codebook](WV6_Codebook.pdf) - detailed description of each variable
- [WV6.rds](../WV6.rds) - complete dataset
- [variable-description-wvs.csv](../variable-description-wvs.csv) - helpful overview and (brief) description of variable names (e.g. for filtering, etc.)

**Participating countries:** 
60 countries (see for details [codebook](../WV6_Codebook.pdf) section *annex* PP002 Country Code p.64)

**Missing data:**
-5 Missing; Unknown
-4 Not asked in survey
-3 Not applicable
-2 No answer
-1 Don´t know

## Overview variable groups
Notice, for the variable groups (SUBSCALE) sometimes a further specification in terms of (SUBSCALE.TOPICS) exists. This wasestablished in order to work easier with the data set. But the assignment of the "topic"-names was partly a bit arbitrary, therefore do not use the SUBSCALE.TOPICS for filtering without criticsm.

- [Perceptions of life](#perceptions-of-life)
- [Environment](#environment)
- [Work](#work) 
- [Family](#family)
- [Politics and Society](#politics-and-society)
- [Religion and Morale](#religion-and-morale)
- [National Identity](#national-identity)
- [Security](#security)
- [Science](#science)
- [Socio-demographics](#socio-demographics)


### load dataset and variable descriptions

```{r dataset-wvs-data}
# read data set
data <- readRDS("WV6.rds")
variables <- read.csv("variable-description-wvs.csv", sep = ";")
str(variables)
```

### Perceptions of life

```{r dataset-wvs-perceptions-of-life}
filter(variables, SUBSCALE == "Perceptions of life") %>%
  select("VARIABLE","LABEL") %>% 
  print()
```

#### SUBSCALE-TOPICS:

```{r dataset-wvs-perceptions-of-life-subscale}
filter(variables, SUBSCALE == "Perceptions of life") %>%
  select("SUBSCALE.TOPICS") %>% 
  distinct() %>% 
  print()
```

### Environment

```{r dataset-wvs-environment}
filter(variables, SUBSCALE == "Environment") %>%
  select("VARIABLE","LABEL") %>% 
  print()
```

#### SUBSCALE-TOPICS:

```{r dataset-wvs-environment-subscale }
filter(variables, SUBSCALE == "Environment") %>%
  select("SUBSCALE.TOPICS") %>% 
  distinct() %>% 
  print()
```

### Work

```{r dataset-wvs-work}
filter(variables, SUBSCALE == "Work") %>%
  select("VARIABLE","LABEL") %>% 
  print()
```

#### SUBSCALE-TOPICS:

```{r dataset-wvs-work-subscale}
filter(variables, SUBSCALE == "Work") %>%
  select("SUBSCALE.TOPICS") %>% 
  distinct() %>% 
  print()
```

### Family

```{r dataset-wvs-family}
filter(variables, SUBSCALE == "Family") %>%
  select("VARIABLE","LABEL") %>% 
  print()
```

#### SUBSCALE-TOPICS:

```{r dataset-wvs-family-subscale}
filter(variables, SUBSCALE == "Family") %>%
  select("SUBSCALE.TOPICS") %>% 
  distinct() %>% 
  print()
```

### Politics and Society

```{r dataset-wvs-politics-society}
filter(variables, SUBSCALE == "Politics and Society") %>%
  select("VARIABLE","LABEL") %>% 
  print()
```

#### SUBSCALE-TOPICS:

```{r dataset-wvs-politics-society-subscale}
filter(variables, SUBSCALE == "Politics and Society") %>%
  select("SUBSCALE.TOPICS") %>% 
  distinct() %>% 
  print()
```

### Religion and Morale

```{r dataset-wvs-religion-morale}
filter(variables, SUBSCALE == "Religion and Morale") %>%
  select("VARIABLE","LABEL") %>% 
  print()
```

#### SUBSCALE-TOPICS:

```{r dataset-wvs-religion-morale-subscale}
filter(variables, SUBSCALE == "Religion and Morale") %>%
  select("SUBSCALE.TOPICS") %>% 
  distinct() %>% 
  print()
```

### National Identity

```{r dataset-wvs-national-identity}
filter(variables, SUBSCALE == "National Identity") %>%
  select("VARIABLE","LABEL") %>% 
  print()
```

#### SUBSCALE-TOPICS:

```{r dataset-wvs-national-identity-subscale}
filter(variables, SUBSCALE == "National Identity") %>%
  select("SUBSCALE.TOPICS") %>% 
  distinct() %>% 
  print()
```

### Security

```{r dataset-wvs-security}
filter(variables, SUBSCALE == "Security") %>%
  select("VARIABLE","LABEL") %>% 
  print()
```

#### SUBSCALE-TOPICS:

```{r dataset-wvs-securtiy-subscale}
filter(variables, SUBSCALE == "Security") %>%
  select("SUBSCALE.TOPICS") %>% 
  distinct() %>% 
  print()
```

### Science

```{r dataset-wvs-science}
filter(variables, SUBSCALE == "Science") %>%
  select("VARIABLE","LABEL") %>% 
  print()
```

#### SUBSCALE-TOPICS:

```{r dataset-wvs-science-subscale}
filter(variables, SUBSCALE == "Science") %>%
  select("SUBSCALE.TOPICS") %>% 
  distinct() %>% 
  print()
```

### Socio-Demographics

```{r dataset-wvs-demographics}
filter(variables, SUBSCALE == "Socio-demographics") %>%
  select("VARIABLE","LABEL") %>% 
  print()
```

#### SUBSCALE-TOPICS:

```{r dataset-wvs-demographics-subscale}
filter(variables, SUBSCALE == "Socio-demographics") %>%
  select("SUBSCALE.TOPICS") %>% 
  distinct() %>% 
  print()
```