## World Values Survey (wave 6 | 2010-2014)

### Nature, origin and rationale of the data

The [World Values Survey](www.worldvaluessurvey.org) (WVS) aims to study *changing values and their impact on social and political life*. The WVS consists of nationally representative surveys conducted in almost 100 countries which contain almost 90 percent of the world's population, using a common questionnaire. The WVS is the largest non-commercial, cross-national, time series investigation of human beliefs and values.

It currently includes interviews with almost *400,000 respondents*. Respondents are people in the age 18 and older residing within private households in each country, regardless of their nationality, citizenship or language.

The main method of data collection in the WVS survey is *face-to-face interview* at respondent's home / place of residence.

#### The questionnaire

The survey was conducted by using a *structured* [questionnaire](../4_WV6_Official_Questionnaire.pdf), consisting of 250 questions (here: variables) ([overview of all variables](../3_WV6_Codebook.pdf)) that describe 10 thematic sub-sections:

1. Perceptions of life,
2. Environment,
3. Work,
4. Family,
5. Politics and Society,
6. Religion and Morale,
7. National Identity,
8. Security,
9. Science, and
10. Socio-demographics.

(The document ["variable description"](../6_variable-description-wvs.csv) contains the assigment of variables to topics.)

#### Theoretical motivation & hypotheses

[**Inglehart's Concept of Postmaterialism**](#postmaterialism-index)
[**Schwartz-Value-Scale**](#value-scale-from-schwartz)
[WORKOUT POSSIBLE HYPOTHESES/ANALYSIS]

### Loading and preprocessing the data

```{r wvs loading data set}
#d_raw_wvs <- readRDS("../data_sets/data-sets/4_world-values-survey/5_WV6.rds")
d_raw_wvs <- readRDS("5_WV6.rds")
head(d_raw_wvs)
```

#### Trust variable

**dummy dependent variable** 
Generally speaking, would you say that most people can be trusted or that you need to be very careful in dealing with people? 

- *trust*(V24)
  + 1 Most people can be trusted.
  + 2 Need to be very careful.

We argue that there are two types of the factors that may shape individual trust intentions. One is denoted by micro-level
factors; the second is linked to macro-level institutions.

**Micro-level factors**

- *Age* (V242) 
- *Gender* (V240)
  + 1 Male
  + 2 Female
- *Education* (V248) 
  + 1 No formal education
  + 2 Incomplete primary school
  + 3 Complete primary school
  + 4 Incomplete secondary school: technical/vocational type
  + 5 Complete secondary school: technical/vocational type
  + 6 Incomplete secondary: university-preparatory type
  + 7 Complete secondary: university-preparatory type
  + 8 Some university-level education, without degree
  + 9 University-level education, with degree
- *Income* (V239)
  + Each country has 10 scales of income: 1-10
- *religious belief** (V144)
No: 
  + 0 do not belong to a denomination 
Yes: 
  + 1 Roman Catholic 
  + 2 Protestant 
  + 3 Orthodox (Russian/Greek/etc.) 
  + 5 Muslim 
  + 99 <-- Other (Reference Group)
- *marital status* (V57) 
  + 1 <-- 1 & 2 Married or living together as married
  + 2 <-- 3 & 4 & 5 & 6 Divorced, Separated, Widowed, Single
- *number of children* (V58)
  + 0 No children
  + 1 One child
  + 2 Two children
  + 3 Three children
  + 4 Four children
  + 5 Five children
  + 6 Six children
  + 7 Seven children
  + 8 Eight or more children

*final religious denominations considered in this study are Muslim, Orthodox, Protestant, Catholic, and "Other". "Other" is taken as the reference group.

```{r wvs-trust select relevant variables}
d_trust <- d_raw_wvs %>% 
  filter(V2 %in% c(276,288,804,156,780)) %>% 
  transmute(
    trust = ifelse(V24==1,1,0),            
    country = factor(V2, levels = c(276,288,804,156,780), labels = c("Germany","Ghana","Ukraine", "China", "Trinidad")),
    age = V242,
    gender = factor(V240, levels = c(1,2), labels = c("male", "female")),
    education = factor(V248, levels = c(1:9), labels = c("no edu.", "incompl. prim. school", "compl. prim. school", "incompl. sec. school-tec", "compl. sec. school-tec", "incompl. sec. school-univ", "compl. sec. school-univ", "univ. without degree", "univ. with degree")),
    income = factor(V239),
    marital = ifelse(V57 %in% c(1,2), "married, together", "single, divorced"),
    children = V58,
    religious_belief = ifelse(V144 == 1, "Catholic",
                              ifelse(V144==2, "Protestant",
                                     ifelse(V144==3, "Orthodox",
                                            ifelse(V144==5, "Muslim",
                                                   "Other"))))
  ) %>% 
  na.omit()

d_trust
```
```{r}
d_trust %>% 
  group_by(country, gender) %>% 
  summarize(
    trust_sum = sum(trust),
    obs = length(trust),
    trust_perc = trust_sum/obs
  )  %>% 
  ggplot(mapping = aes(x = country, y = trust_perc, fill = gender)) +
  geom_col(position = "dodge") 
```
```{r}
d_trust %>% 
  group_by(country, income, marital) %>% 
  summarize(
    trust_sum = sum(trust),
    obs = length(trust),
    trust_perc = trust_sum/obs
  )  %>% 
  ggplot(mapping = aes(x = income, y = trust_perc, fill = marital)) +
  geom_col(position = "dodge") +
  facet_wrap(~country, nrow = 1)
```


```{r}
d_trust_DE <- d_trust %>% 
  filter(country=="Germany")
```


```{r}
model1 = trust ~ age + gender + education + income + marital + children

brm(formula = model1, data = d_trust_DE, family = "bernoulli")
```



