## World Values Survey (wave 6 | 2010-2014)

### Nature, origin and rationale of the data

The [World Values Survey](www.worldvaluessurvey.org) (WVS) aims to study *changing values and their impact on social and political life*. The WVS consists of nationally representative surveys conducted in almost 100 countries which contain almost 90 percent of the world's population, using a common questionnaire. The WVS is the largest non-commercial, cross-national, time series investigation of human beliefs and values.

It currently includes interviews with almost *400,000 respondents*. Respondents are people in the age 18 and older residing within private households in each country, regardless of their nationality, citizenship or language.

The main method of data collection in the WVS survey is *face-to-face interview* at respondent's home / place of residence.

#### The questionnaire

The survey was conducted by using a *structured* [questionnaire](../4_WV6_Official_Questionnaire.pdf), consisting of 250 questions (here: variables) ([overview of all variables](../3_WV6_Codebook.pdf)) that describe 10 thematic sub-sections:

1. Perceptions of life,
2. Environment,
3. Work,
4. Family,
5. Politics and Society,
6. Religion and Morale,
7. National Identity,
8. Security,
9. Science, and
10. Socio-demographics.

(The document ["variable description"](../6_variable-description-wvs.csv) contains the assigment of variables to topics.)

### Loading and preprocessing the data

```{r wvs loading data set}
#d_raw_wvs <- readRDS("../data_sets/data-sets/4_world-values-survey/5_WV6.rds")
d_raw_wvs <- readRDS("5_WV6.rds")
head(d_raw_wvs)
```

#### Case story: How Stable Are Democracies? 'Warning Signs Are Flashing Red' 

The following example is taken from Healy (2018) [INCLUDE REFERENCE]. 

The descriptive analysis and plot is reproduced from Foa and Mounk (2016), with the following changes: Foa and Mounk (2016) used combined data from Waves 5 and 6 (2005-7 and 2010-14) in the current analysis only data from wave 6 are used. Therefore one country "Great Britain" doped out. It has been replaced with "Germany".

**relevant variables**
- *country* (V2)
  + 36 Australia
  + 276 Germany (as replacement for Great Britain (826); was not asked in wave 2010-2014)
  + 528 Netherlands
  + 554 New Zealand
  + 752 Sweden
  + 840 Unitated States
- *year of birth* (V241)
  + 30th (30-39)
  + 40th (40-49)
  + 50th (50-59)
  + 60th (60-69)
  + 70th (70-79)
  + 80th (80-89)
- *How important is it for you to live in a country that is governed democratically?* (V140)
  + rating scale: 1-"not at all important" to 10-"Absolutely important"

```{r wvs-democracy selecting relevant variables}
d_democracy <- d_raw_wvs %>% 
  select(V2,V241,V140) %>% 
  filter(V2 %in% c(36,276,528,554,752,840) & V241 >= 1930 & V241 < 1990) %>% 
  mutate(
    birth_decade = ifelse(V241 >= 1930 & V241< 1940, 1930,
                          ifelse(V241 >= 1940 & V241< 1950, 1940,
                                 ifelse(V241 >= 1950 & V241< 1960, 1950,
                                        ifelse(V241 >= 1960 & V241< 1970, 1960,
                                               ifelse(V241 >= 1970 & V241 < 1980, 1970,
                                                      1980))))),
    democracy_essential = ifelse(V140 == 10,1,0)
  ) %>% 
  replace_with_na_all(condition = ~.x %in% c(-5,-4,-3,-2,-1)) %>% 
  mutate(
    country = factor(V2, levels = c(36,276,528,554,752,840), labels = c("Australia","Germany","Netherlands","New Zealand","Sweden","United States")),
    rating = factor(V140, ordered = TRUE)
    ) %>% 
  na.omit() %>% 
  print()
```


```{r wvs-democracy create percentage of rating=10 (dichotomous) for country and year}
d_democracy_grouped <- d_democracy %>% 
  group_by(birth_decade, country) %>% 
  summarise(
    democracy = sum(democracy_essential),
    obs = length(V140),
    democracy_per = democracy/obs,
    lower_ci = mean(democracy_essential)-2*sd(democracy_essential)/sqrt(length(democracy_essential)),
    upper_ci = mean(democracy_essential)+2*sd(democracy_essential)/sqrt(length(democracy_essential))
  ) %>% 
  print()
```


```{r wvs-democracy plotting democracy-attitude per birth year and country , fig.cap="Reproducing plot by Foa and Mounk (2016): democracy-attitude as dichotomous variable across age and countries"}
d_democracy_grouped %>% 
  ggplot(mapping = aes(x = birth_decade, y = democracy_per))+
  geom_point() +
  geom_line() +
  geom_ribbon(mapping = aes(x = birth_decade, ymin = lower_ci, ymax = upper_ci), alpha = .2) +
  facet_wrap(~ country, nrow = 1) +
  theme(axis.text.x = element_text(angle = 90)) +
  scale_x_continuous(name = "Decade of Birth", breaks = c(1930,1940,1950,1960,1970,1980)) +
  scale_y_continuous(name = ""Essential" to Live in a Democracy", limits = c(.15,.9), breaks = c(0.2,0.4,0.6,0.8), labels = c("20%","40%","60%","80%"))

```

Healy (2016) cites Eric Voeten who redraw the above plot by using the 10-point rating variable instead of reduced dichotomous variant of the dependent variable. In the following a reproduction of Voeten's graphic:

```{r wvs-democracy create mean for 10-rating-scale per country and birth year}
d_democracy_grouped2 <- d_democracy %>% 
  group_by(V241, country) %>% 
  summarise(
    democracy = mean(V140),
    lower_ci = mean(V140)-2*sd(V140)/sqrt(length(V140)),
    upper_ci = mean(V140)+2*sd(V140)/sqrt(length(V140))
  ) 
```

```{r wvs-democracy plotting democracy-attitude per birth year and country , fig.cap="Reproducing plot by Eric Voeten [CITATION FROM HEALY-BOOK]: democracy-attitude as 10-rating scaled variable across age and countries"}
d_democracy_grouped2 %>%   
  ggplot(mapping = aes(x = V241, y = democracy))+
  geom_smooth(color = "black") +
  facet_wrap(~ country, nrow = 1) +
  scale_x_continuous(name = "Decade of Birth", breaks = c(1930,1950,1970,1989), labels = c("'30", "'50", "'70", "'90"), limits = c(1930,1990)) +
  scale_y_continuous(name = ""Essential" to Live in a Democracy", limits = c(1,10),breaks = c(1,2,3,4,5,6,7,8,9,10))
```

After visualizing the data some analysis can be made in order to check if the usage of a dichotomous or 10-point dependent variable influence the adequacy of a model.

```{r wvs-democracy one predictor-model (birth year) for country "Australia"}
data <- d_democracy %>% 
  filter(country =="Australia")

model1 <- democracy_essential ~ V241 
model2 <- rating ~ V241 

m1 <- brm(data = data, formula = model1, family = "bernoulli")
m2 <- brm(data = data, formula = model2, family = "cumulative")
```

```{r wvs-democracy model fit summary: rating dichotomous; one predictor-model}
summary(m1)
```


```{r wvs-democracy model fit summary: rating 10-scaled; one predictor-model}
summary(m2)
```

```{r}
m1 <- add_criterion(m1, "waic")
m2 <- add_criterion(m2, "waic")

print(m1$waic)
print(m2$waic)

loo_compare(m1,m2, criterion = "waic")
```

```{r wvs-democracy model fit plot: rating dichotomous; one predictor-model, fig.cap="Predictions of the dichotomous model"}
# model predictions 
plot(marginal_effects(m1), points = TRUE, rug = TRUE)
```

```{r wvs-democracy predictions vs. observed data: rating dichotomous; one predictor-model}
# Predicted responses based on model fit
pp_m1 <- predict(m1)
pred <- pp_m1[,1]

# include predictions in observed data set
data <- data %>% 
  add_column(
    pred = as_vector(pred)
  ) %>% 
  mutate(
    pred_m1 = ifelse(pred < 0.5, 0,1)
  ) 
# create crosstable (Package: gmodels)
CrossTable(data$pred_m1,data$democracy_essential, prop.r = FALSE, prop.c = FALSE, prop.chisq = FALSE, dnn = c("observed data","predicted data"))
```

```{r wvs-democracy model fit plot: rating 10-scaled; one predictor-model, fig.cap="Predictions of the 10-rating scale model"}
# model predictions
marginal_effects(m2, categorical = TRUE)
```

```{r wvs-democracy predictions vs. observed data: rating 10-scaled; one predictor-model}
# Predicted responses based on model fit
pp_m2 <- predict(m2) %>% 
  as_tibble %>% 
  `colnames<-`(c(1,2,3,4,5,6,7,8,9,10))

pred_m2 <- colnames(pp_m2)[apply(pp_m2,1,which.max)]

# include predictions in observed data set
data <- data %>% 
  add_column(
    pred = as_vector(pred_m2)
  ) 
# create crosstable (Package: gmodels)
CrossTable(data$pred,data$rating, prop.r = FALSE, prop.c = FALSE, prop.chisq = FALSE, dnn = c("observed data","predicted data"))
```


Until here there some problems with estimation, perhaps due to false model? 

```{r wvs-democracy group-level model; group: country("Australia", "New Zealand")}
data2 <- d_democracy %>% 
  filter(country %in% c("Australia", "New Zealand"))

model21 <- democracy_essential ~ V241 + V241|country 
model22 <- rating ~ V241 + V241|country

m21 <- brm(data = data2, formula = model21, family = "bernoulli")
m22 <- brm(data = data2, formula = model22, family = "cumulative")

summary(m21)
```

```{r}
summary(m22)
```

```{r}
# model predictions 
plot(marginal_effects(m21), points = TRUE, rug = TRUE)
```

```{r}
# Predicted responses based on model fit
pp_m21 <- predict(m21)
pred <- pp_m21[,1]

# include predictions in observed data set
data <- data %>% 
  add_column(
    pred = as_vector(pred)
  ) %>% 
  mutate(
    pred_m1 = ifelse(pred < 0.5, 0,1)
  ) 
# create crosstable (Package: gmodels)
CrossTable(data$pred_m1,data$democracy_essential, prop.r = FALSE, prop.c = FALSE, prop.chisq = FALSE, dnn = c("observed data","predicted data"))
```

```{r}
# model predictions
marginal_effects(m22, categorical = TRUE)
```

```{r}
# Predicted responses based on model fit
pp_m22 <- predict(m22) %>% 
  as_tibble %>% 
  `colnames<-`(c(1,2,3,4,5,6,7,8,9,10))

pred_m22 <- colnames(pp_m2)[apply(pp_m22,1,which.max)]

# include predictions in observed data set
data <- data %>% 
  add_column(
    pred = as_vector(pred_m22)
  ) 
# create crosstable (Package: gmodels)
CrossTable(data$pred,data$rating, prop.r = FALSE, prop.c = FALSE, prop.chisq = FALSE, dnn = c("observed data","predicted data"))
```

