
# General description

The European Social Survey (ESS) is an academically driven cross-national survey that has been conducted across Europe since its establishment in 2001. Every two years, face-to-face interviews are conducted with newly selected, cross-sectional samples. The survey measures the attitudes, beliefs and behaviour patterns of diverse populations in more than thirty nations.

**Actual version:** 
ESS round 8, year: 2016

**Further information about ESS:** 
https://www.europeansocialsurvey.org/about/ 

**Citation of downloaded data set:** 
ESS Round 8: European Social Survey Round 8 Data (2016). Data file edition 2.1. NSD - Norwegian Centre for Research Data, Norway - Data Archive and distributor of ESS data for ESS ERIC. doi:10.21338/NSD-ESS8-2016.

**Data collection mode:**
structured questionnaires

# Data set - description

**Detailed description of current data set and each variable:**
See [ESS8.pdf](../ESS8.pdf) in folder *european-social-survey*

**Year:** 
2016

**Number of countries:** 
23 countries (for details see section [Countries](#countries))

**Missing data:**

- 6, 66 etc. = Not applicable
- 7, 77 etc. = Refusal
- 8, 88 etc. = Don't know
- 9, 99 etc. = No answer

## Overview variable groups

 - [Media and social trust](#media-and-social-trust) (6 variables)
 - [Politics](#politics) (91 variabels)
 - [Subjective well-being, social exclusion, religion, national and ethnic identity](#subjective-well-being-social-exclusion-religion-national-and-ethnic-identity) (75 variables)
 - [Climate change](#climate-change) (32 variables)
 - [Welfare attitudes](#welfare-attitudes) (42 variables)
 - [Gender, Year of birth and Household grid](#gender-year-of-birth-and-household-grid) (37 variabels)
 - [Socio-demographics](#socio-demographics) (210 variables)
 - [Human values](#human-values) (21 variables)
 
### load dataset and variable descriptions

```{r dataset-ess-data}
# read data set
data <- read.csv("ESS8.csv")
# read variable descriptions
variables <- read.csv("variables.csv")
```

### Countries 

```{r dataset-ess-countries}
# countries
print(levels(data$cntry))
```

### Media and social trust 

```{r dataset-ess-media-social-trust}
filter(variables, Group=="Group Media and social trust") %>% 
  print()
```

### Politics

```{r dataset-ess-politics}
filter(variables, Group=="Group Politics") %>% 
  print()
```

### Subjective well-being, social exclusion, religion, national and ethnic identity

```{r dataset-ess-subjective-well-being}
filter(variables, Group=="Group Subjective well-being, social exclusion, religion, national and ethnic identity") %>% 
  print()
```

### Climate change

```{r dataset-ess-climate-change}
filter(variables, Group=="Group Climate change") %>% 
  print()
```

### Welfare attitudes

```{r dataset-ess-welfare-attitudes}
filter(variables, Group=="Group Welfare attitudes") %>% 
  print()
```

### Gender, Year of birth and Household grid

```{r dataset-ess-gender}
filter(variables, Group=="Group Gender, Year of birth and Household grid") %>% 
  print()
```

### Socio-demographics

```{r dataset-ess-demographics}
filter(variables, Group=="Group Socio-demographics") %>% 
  print()
```

### Human values

```{r dataset-ess-human-values}
filter(variables, Group=="Group Human values") %>% 
  print()
```
