# Basics of R

```{block2, type='overview'}
**Content covered:** TBF

**Learning goal:** familiarize yourself with the basic of R and the tidyverse
```

## General remarks

- R is specialized programming language for statistics/data science
- base R vs tidyverse
- packages, CRAN, github
- resources:
  - official (base) R intro: [An Introduction to R](https://colinfay.me/intro-to-r/)
  - book on DS with tidyverse: [R for Data Science (R4DS)](https://r4ds.had.co.nz)

## Installation and use

- installation of R itself
- RStudio
- how to install packages
- which packages to install for this course
  - tidyverse, brms, ...

## First steps

- scripts and sessions
  - interpreted language
- using RStudio
- calculations, variable assignments 
  - good naming conventions
- comments
  - sectioning in RStudio
  - good practices for literate programming
- everything is vector/matrix based
  - declare vector with `c()`
  - indexing starting from 1
  - matrices in "column-major mode" (unlike NumPy!) -> first index moves fastest
```{r}
(m = matrix(1:6, ncol = 3))
```
  - matrix indexing starts with row
```{r}
m[1,]
```
  - vectors, therefore, are column vectors:
```{r}
x = c(1,0,1)
m %*% x
```
- getting help / information about function / vignettes
- 


## Data types

### Numbers

- standard number precision is double

### Booleans

### Special values

- NA, NaN, Inf, ... what else?

### Characters (= strings)

- strings are called characters

```{r}
typeof("huhu")
```

- `paste`, `grep`, `cat`, `sprintf`... what else?

- vector of characters

```{r}
chr.vector = c("huhu", "hello", "huhu", "ciao")
chr.vector
```

### Factors

- factors track levels

```{r}
factor(chr.vector)
```

- ordered factors arrange their levels

```{r}
factor(chr.vector, ordered = T, 
       levels = c("huhu", "ciao", "hello"))
``` 

### Lists, data frames & tibbles

- lists are key-value pairs

```{r}
my_list = list(dudu = 1,
               chacha = c("huhu", "ciao"))
```

- data frames as lists of same-length vectors

```{r}
exp_data = data.frame(
  trial = 1:5,
  condition = factor(c("C1", "C2", "C1", "C3", "C2"),
                     ordered = T),
              response = c(121, 133, 119, 102, 156))
exp_data
```

- access colums

```{r}
exp_data$condition
```

- access rows

```{r}
exp_data[3,]
```

- tibbles are data frames in the tidyverse

```{r}
as_tibble(exp_data)
```

- tibbles do but data frames don't allow for dynamic construction:

- some differences

```{r, eval = FALSE}
my_tibble    = tibble(x = 1:10, y = x^2)      ## dynamic construction possible
my_dataframe = data.frame(x = 1:10, y = x^2)  ## ERROR :/
```

## Functions

- named and anonymous

## Loops and maps

- `for` loops
- functional programming with `apply` `purrr`
  - the philosophy of "passing functions to functions"
- piping