# (PART) Data {-}

# Data, variables & experimental designs {#Chap-02-01-data}

The focus of this course is on data from behavioral psychology experiments.^[A *behavioral experiment* is an experiment that records participants' behavioral choices, such as button clicks or linguistic responses in the form of text or speech. This contrasts with, say, *neurological experiments* in which participants' brain activity is recorded, such as fMRI or EEG, or, e.g., in a psycholinguistic context, *processing-related experiments* in which secondary measures of cognitive activity are measured, such as eye-movements, pupil dilation or galvanic skin responses.] In a sense, this is perhaps the most "well-behaved" data to analyze and therefore an excellent starting point into data analysis. However, we should not lose sight of the rich and diverse guises of data that are relevant for scientific purposes. The current chapter therefore starts, in Section \@ref(Chap-02-01-data-kinds-of-data) with sketching some of that richness and diversity. But it then hones in on some basic distinctions of the kinds of data we will frequently deal with in the cognitive sciences in Section \@ref(Chap-02-01-data-variables). We also pick up a few relevant concepts from standard experimental design in Section \@ref(Chap-02-01-data-exp-design) and learn about what makes data tidy in Section \@(Chap-02-01-data-tidy-data).

<!-- TODO: insert references to subsections -->

```{block, type='infobox'}
The learning goals for this chapter are:

- appreciate the diversity of data
- distinguish different kinds of variables
  - dependent vs independent
  - nominal vs ordinal vs metric
- get familiar with basic aspects of experimental design
  - factorial designs
  - within- vs between subjects design
  - repeated measures
  - randomization, fillers and controls
  - sample size
- understand the notion of "tidy data"
```

## Different kinds of data {#Chap-02-01-data-kinds-of-data}

Some say we live in the **data age**. But what is data actually? Purist pedants say: "The plural of datum" and add that a datum is just an observation. But when we say "data" we usually mean a bit more than a bunch of observations. The Merriam-Webster offers the following:

> Factual information (such as measurements or statistics) used as a basis for reasoning, discussion, or calculation.

This is a teleological definition in the sense that it refers to the purpose of the thing: "used as basis for reasoning, discussion, or calculation". This is important, "data" is defined by what we do with it. Another important aspect of this definition is fairly backgrounded, but just as important. We usually consider data to be systematically structured in some way or another. Even when we speak of "raw data", we expect there to be some structure (maybe labels, categories etc.) that distinguishes data from uninterpretable noise (see below the notion of a "variable", discussed in Section \@ref(Chap-02-01-data-variables)). In sum, we can think of data as a represenation of information, stored in a systematic way for the purpose of inference, argument or decision making.

It is easy but wrong to think that data always has to be information based on observations of the world. It is easy to think this because **empirical data**, i.e., data obtained from empirical observation, is the most common form of data (given that it is, arguably, most relevant for decision making and argument). But it is wrong to think this because we can just as well look at **virtual data**. For example, *virtual data* could be **data from computer simulation studies**, e.g., from, say, a 1 billion runs of a multi-agent simulation intended to shed light on the nature of cooperative interaction. It makes sense to analyse such data with the same tools as data from an experiment. We might find out that some parameter constellations in the simulation run are (statistically) most conducive of producing cooperative behavior among our agents, for instance. Another example of *virtual data* is so-called **repeat data** or **fake data** generated from a statistical model, which we can use to test whether that model is any good, in so-called model criticism (see Section \@ref(Chap-03-08-model-criticism)).^[We will later speak of **prior/posterior predictions** for this kind of repeat data.] Finally, **hypothetical data** is data that was neither observed, nor predicted by a model, but something that could have been observed, something that it is merely logically possible to observe, even if it would almost never happen in reality or would not be predicted by any serious model.

The most frequent form of data, **empirical data** about the actual world, comes in two major variants. **Observational data** is data gathered by (passively) observing and recording what would have happened anyone, so to speak. Examples of observational data are collections of socio-economic variables, like gender, education, income, number of children etc. In contrast, **experimental** data is data recorded in a strict regime of manipulation-and-observation, i.e., a scientific experiment. Some pieces of information can only be recorded in an observation study (annual income) and others can only be obtained through experimentation (memory span). Both methods of data acquisition have their own pros and cons. Here are some of the more salient ones:


```{r, echo = F}
table_data <- tribble(
  ~observational, ~experimental,
  "ecologicall valid", "possibly artificial",
  "easy/easier to obtain", "hard/harder to obtain",
  "correlation & causation hard to tease apart", "may yield information on causation vs. correlation"
)
knitr::kable(
  table_data,
  caption = "Comparison of pro's and cons's of observational data and experimental data.", 
  booktabs = TRUE
)
```


No matter what kind of data we have at hand, there at least two prominent purposes for which data can be useful: **explanation** and **prediction**. Though related, it is useful to keep these purposes cleanly apart. Data analysis for explanation uses the data to better understand the source of the data (the world, a computer simulation, a model, etc.). Data analysis for prediction tries to extract regularities from the data gathered so far to make predictions about future data that are as accurate as possible.

## On the notion of "variables" {#Chap-02-01-data-variables}

Data used for data analysis, even if it is "raw data", i.e., data before preprocessing and cleaning, is usually structured or labelled in some way or other. Even if the whole data we have is a vector of numbers, we would usually know what these numbers represent. For instance, we might just have a quintuple of IQ scores or a Boolean vector of whether five students passed the exam; we then still (usually) know what these data points represent:

```{r}
# a simple data vector of IQ-scores
IQ_scores <- c(102, 115, 97, 126, 87)

# who passed the exam
exam_results <- 
  tribble(
    ~student,   ~pass,
    "Jax",      TRUE,
    "Jason",    FALSE,
    "Jamie",    TRUE
  )
```

Most often, however, we have more than one kind of observation that we care about. Most often, we care about systematic relationships and so we care about co-occurrences of observations of different kind. For instance, we might want to know look at pass/fail results from exam in co-occurrence with information about who attended tutorials:

```{r}
# proportion of tutorials attended and exam pass/fail
exam_results <- 
  tribble(
    ~student,   ~tutorial_proportion,   ~pass,
    "Jax",      0.0,                    TRUE,
    "Jason",    0.78,                   FALSE,
    "Jamie",    0.39,                   TRUE
  )
exam_results
```

Data of this kind is also called **rectangular data**, i.e., data that fits into a rectangular table.(More on the structure of rectangular data in Section \@ref(Chap-02-01-data-tidy-data).) In the example above, every column represents a **variable** of interest. A *(data) variable* stores the observations that are of the same kind.^[This sense of "data variable" is not to be confused with the notion of a "random variable", a concept we will introduce later in Section \@(Chap-03-01-probability-random-variables). The term "data variable" is not commonly used; it is merely "variable".]

Common kinds of variables are distinguished based on the structural properties of the kinds of observations that they contain. Common types of empirical data collected are:

- **nominal variable**: each observation is an instance of a (finite) set of clearly distinct categories, lacking a natural ordering
- **binary variable**: special case of nominal variable when there are only two categories
- **Boolean variable**: special case of a binary variable when the two categories are Boolean values "true" and "false"
- **ordinal variable**: each observation is an instance of a (finite) set of clearly distinct and naturally ordered categories, but there is no natural meaning of distance between categories (i.e., it makes sense to say that A is "more" than B but not that A is three times "more" than B)
- **metric variable**: each observation is isomorphic to a subset of the reals, and interval-scaled (i.e., it makes sense to say that A is three times "more" than B)

<!-- TODO: add examples -->

In experimental data we also distinguish the **dependent variable(s)** from the **independent variables**. The dependent variables are the variables that we do not control or manipulate in the experiment, but the ones that we are curious to record (e.g., whether a patient recovered from an illness within a week). Dependent variables are also called **to-be-explained variables**. The independent variables are the variables in the experiment that we manipulate (e.g., which drug to administer), usually with the intention of seeing a particular effect on the dependent variables. Independent variables are also called **explanatory variables**.

## Basics of experimental design {#Chap-02-01-data-exp-design}

- open/closed questions
- factorial designs
- within- vs between-subjects design
- repeated measures
- randomization, fillers and controls
- sample size
  
## Tidy data {#Chap-02-01-data-tidy-data}

- stages: preprocesssing & cleaning
- wrangling
- long, wide & tidy formats

