# Data

*learning goal:* how to arrange, summarize and visualize (aspects of data) to address a question of interest ("hypothesis-driven data poking")



- different kinds of data
- summary statistics
- data wrangling
- data plotting

## Kinds of data


## Summary statistics


### Exploring numerical data in a vector

The first part of the Mental Chronometry experiment is a simple reaction time task.

```{r}
# read a data set of reaction times (as a tibble)
RTs <- read_csv("https://tinyurl.com/y4jo8mox")

# check its content
glimpse(RTs)
```

## Data Wrangling

## Data Plotting

Numerical summaries of complex data always incur information loss. Still lossy, but less so is visualization. Any serious data analysis should start with a process in which the analyst becomes intimate with the data at hand. Visualization is an integral part of data-intimacy.

Consider a famous dataset available in R [@anscombe1973]: 

```{r}
glimpse(anscombe %>% as_tibble)
```

There are four pairs of $x$ and $y$ coordinates. Unfortunately, these are stored in long format with two pieces of information buried inside of the column name: $x3$ contains the information that this column contains the $x$ coordinates for the 3rd pair. This is rather untidy. But, after the last section on data wrangling, we can tidy up quickly:

```{r}
tidy_anscombe = anscombe %>% as_tibble %>% 
  pivot_longer(
    1:8,                          # columns to pivot
    names_sep = 1,                 # position indicating where to split each column name
    names_to = c("axis", "group")  # names for new columns
    ) 
tidy_anscombe
```

Here are some summary statistics for each of the four pairs:

```{r}
tidy_anscombe %>% 
  group_by("group") %>% 
  summarise()
```



This is a plot of quite a famous dataset [@anscombe1973]:

```{r ch-02-anscombe, fig.cap="Anscombe's Quartet: four different data sets all of which receive the same correlation score."}
tibble(
  grp = rep(c("I", "II", "III", "IV"), each = 11),
  x = c(anscombe$x1, anscombe$x2, anscombe$x3, anscombe$x4),
  y = c(anscombe$y1, anscombe$y2, anscombe$y3, anscombe$y4)
) %>% 
  ggplot(aes(x, y)) +
    geom_smooth(method = lm, se = F, color = "black") +
    geom_point(color = project_colors[3], size = 2) +
    scale_y_continuous(breaks = scales::pretty_breaks()) +
    scale_x_continuous(breaks = scales::pretty_breaks()) +
    labs(title = "Anscombe's Quartet", x = NULL, y = NULL,
         subtitle = bquote(y == 0.5 * x + 3 ~ (R^2 %~~% .667) ~ "for all datasets")) +
    facet_wrap(~grp, ncol = 2, scales = "free_x") +
    theme(strip.background = element_rect(fill = "#f2f2f2", colour = "white"))
```
