
# Beyond simple linear regression {#Chap-04-02-beyond-simple-regression}

<hr>

This chapter showcases how linear regression models can be applied flexibly to a variety of data analysis problems.

- categorical predictors
  - two-groups (eco-sensitivity data)
  - more groups (mental chronometry)
  - interactions (Winter data)
- metric and categorical predictors combined
  - avocado data
- logistic regression
  - KoF data


## Two categorical predictors

Let's revisit the (fictitious) eco-sensitivity data from Chapter \@ref(ch-03-07-hypothesis-testing-Bayes):

```{r}
x_A <- c(
  104, 105, 100, 91, 105, 118, 164, 168, 111, 107, 136, 149, 104, 114, 107, 95, 
  83, 114, 171, 176, 117, 107, 108, 107, 119, 126, 105, 119, 107, 131
)
x_B <- c(
  133, 115, 84, 79, 127, 103, 109, 128, 127, 107, 94, 95, 90, 118, 124, 108, 
  87, 111, 96, 89, 106, 121, 99, 86, 115, 136, 114
)
```

Remember that we are interested in the question whether there is a difference in means of group A and group B. Previously we used a $t$-test for this two-group comparison (frequentist or Bayesian). But we can also cast this as a regression problem. Here is how.

Let's first squeeze the data into a tibble:

```{r}
eco_sensitivity_data <- tibble(
  group = c(rep("A", length(x_A)), rep("B", length(x_B))),
  measurement = c(x_A, x_B)
) 
eco_sensitivity_data
```

Notice that this tibble contains the data in a tidy format, i.e., each row contains a tuple of associated measurements. We want to explain or predict the variable `measurement` in terms of the variable `group`.We can then run a regression model with forumla `measurement ~ group`. Here's such a model using the Bayesian approach:

```{r, eval = F}
fit_brms_eco_sensitivity <- brm(
  # specify what to explain in terms of what
  #  using the formula syntax
  formula = measurement ~ group,
  # which data to use
  data = eco_sensitivity_data
)
```

```{r, echo = F}
fit_brms_eco_sensitivity <- readRDS('models_brms/fit_brms_eco_sensitivity.rds')
```

Let's inspect the summary information for the posterior samples:

```{r}
# just showing information for the so-called 'fixed effects'
summary(fit_brms_eco_sensitivity)$fixed 
```

Compare this with the summary of the posterior estimates we obtained from the Bayesian $t$-test model for this data which we implemented in `greta` and ran in Chapter \@ref(ch-03-07-hypothesis-testing-Bayes).

```{r echo = F, eval = T}
draws_t_test_2 <- readRDS('models_greta/ttest_2_draws.rds')
tidy_draws_tt2 = ggmcmc::ggs(draws_t_test_2)
# get means and 95% HDI
Bayes_estimates_eco <- tidy_draws_tt2 %>% 
  group_by(Parameter) %>%
  summarise(
    '|95%' = HDInterval::hdi(value)[1],
    mean = mean(value),
    '95|%' = HDInterval::hdi(value)[2]
  )
```

```{r}
Bayes_estimates_eco
```

The mean of the $\delta$ parameter looks suscpiciously similar to the ominous `groupB` parameter shown in the output of the `bmrs` model fit. The 95% HDIs of the estimated posterior for the $\delta$ parameter also look suspiciously like the values of the 95% interquantile range for the `groupB` parameter. This is no conincidence! In fact, the regression model that `bmrs` calculates here is essentially the same as the $t$-test model we implemented in `greta` by hand except for slight different in the choice of the priors and inessential differences in mathematical formulation of the group mean comparison. 

- show plots of posterior side-by-side
- spell out models, show graphs

The upshot is, that we can conceive of a $t$-test as a special case of a linear regression model!
