# Solutions to exercises {#app-98-solutions-exercises}
<div class = "exercises">
**Exercise E.1: Some algebra**

Solve for $x$: 

a. $3x = 3$
b. $4x = 2$

<div class="collapsibleSolution">
<button class="trigger">Solution</button>
<div class="content">
a. $x = 1$
b. $x = 0.5$
</div>
</div>
</div>

<!-- exercise 2 -->
<div class = "exercises">
**Exercise E.2: Indexing**

Predict the outcome of this:

```{r, eval = T}
x <- c(1,2,3,4)
x[1]
```

<div class="collapsibleSolution">
<button class="trigger">Solution</button>
<div class="content">

We expect `x = 1`. Let's try:

```{r}
x[1]
```
  
</div>
</div>

</div>


Solutions to exercises are provided in the following.

## Chapter 1
<div class = "exercises">
**Exercise E.1**

Calculate in R concole (659-34)*5 
<div class="collapsibleSolution">
<button class="trigger">Solution</button>
<div class="content">

3125

</div>
</div>
</div>


<div class = "exercises">

**Exercise E.2**

create 2 variables and assign values 103 and 14 to them. Next, devide first one by the second and produce an output with 3 digits after comma.

<div class="collapsibleSolution">
<button class="trigger">Solution</button>
<div class="content">
7.357 
</div>
</div>
</div>

<div class = "exercises">
**Exercise E.3**

Create a vector wich contaions all even numbers drom 0 to 20 and assign it to a variable.  Now transform the variable such that it contains only odd numbers up to 20 using mathematical operation. Notice that the numbers avove 20 should not be included! HINT: you may use indexing to fullfill this creteria.

<div class="collapsibleSolution">
<button class="trigger">Solution</button>
<div class="content">
```{r}
a <- seq(from = 0, to = 20, by = 2) 
a <- a+1
a=a[1:10]
a
```
</div>
</div>
</div>

<div class = "exercises">
**Exercise E.4**

Create a matrix with three columns and three rows, wich contains random integers from 1 to 100 without replacement. HINT! You may first produce an array with random integers and then parse it to the matrix() function

<div class="collapsibleSolution">
<button class="trigger">Solution</button>
<div class="content">
```{r}
a = sample(1:100, 9, replace=F)
b= matrix(a, ncol=3)
b
```

</div>
</div>
</div>

<div class = "exercises">
**Exercise E.5**

Ex. Create a vector "a" wich contains names of 3 your best friends and a vector "b" with their age. Create a data frame, wich represents this information (One column with names ant one with respective age). Notice that column names should represent the information they contain!
<div class="collapsibleSolution">
<button class="trigger">Solution</button>
<div class="content">
```{r}
a= c("M","N","H")
b=c(23,41,13)
best_friends <- data.frame(name=a,age=b)
```
</div>
</div>
</div>

<div class = "exercises">
**Exercise E.6**

Print out only names of friends from best_friends data frame.
<div class="collapsibleSolution">
<button class="trigger">Solution</button>
<div class="content">

```{r}
best_friends["name"] 
best_friends[1] 
```
</div>
</div>
</div>

<div class = "exercises">
**Exercise E.7**

Print out only names of friends who are younger then 22.
 
<div class="collapsibleSolution">
<button class="trigger">Solution</button>
<div class="content">
```{r}
which(best_friends["name"]<=22)
```
</div>
</div>
</div>


<div class = "exercises">
**Exercise E.8**

Create a vector "a" wich contains names of 3 your best friends and a vector "b" with their age. Create a tibble with two colums: name, age. Print it. Now, add a new column named "birht date" and the respective values. Add another column, which states TRUE,if the date of birth is even. (HINT: you can create a new column dynamicly! First you have to declare a column on the left sight (tibble["name-of-the-column"]) , and then assign a value to it. For the latter you might check out modulo operator (%%).
<div class="collapsibleSolution">
<button class="trigger">Solution</button>
<div class="content">
```{r}
a= c("M","N","H")
b=c(23,41,13)
best_friends <- tibble(name=a,age=b)
best_friends["birht date"] = c(3, 5,6)
best_friends["is even"] = best_friends["birht date"]%%2==0
best_friends

```
</div>
</div>
</div>

<div class = "exercises">
**Exercise E.9**


Create a function, which takes two numbers and outputs 0 if their product is less than 100, and 1 otherwise. 

<div class="collapsibleSolution">
<button class="trigger">Solution</button>
<div class="content">
```{r}

bigger_100 <- function(x,y) {
 
  return(as.integer( x*y>100))
}
```
</div>
</div>
</div>

<div class = "exercises">
**Exercise E.10**

How many variables should you parse to a function if:

a. it tells if a sum of two numbers even?

b. it applies 2 different operations to a variable and sums the results. Operations are not fixed in the function. 
<div class="collapsibleSolution">
<button class="trigger">Solution</button>
<div class="content">

a. 2 arguments

b. 3 arguments

</div>
</div>
</div>

<div class = "exercises">
**Exercise E.11**

Let's practice for-loops and if/else statments!
Create a vector A with 10 random integers in range (1:50) and the other. Then, using for-loop make the second, filled with the mean of the corresponding intenger in vector A, one integer above it, and one after (For first and last intenger missing values are equal to 0). Print out the result as a tibble. 

<div class="collapsibleSolution">
<button class="trigger">Solution</button>
<div class="content">
```{r}

A=c(sample((1:50), 10))
B=c(integer(length(A)))

    for (i in 1:length(A)){
       if (i  == 1) {
         B[i] <- (sum(A[i:(i+1)])/3)
       }
        else if (i  == length(A)) {
         
         B[i] <-(sum((A[(i-1):i]))/3)
       }
         
        else {
         B[i] <- (mean(A[(i-1):(i+1)] ))
       }
    }
result <-tibble(A,B)

```

</div>
</div>
</div>


## Chapter 8
<div class = "exercises">
**Exercise E.1**   

 Calculate the probability of observing no more than 7 successes for $N= 24$ and $\theta =0.2$.
 
<div class="collapsibleSolution">
<button class="trigger">Solution</button>
<div class="content">


```{r, echo = F}
  pbinom(7, 24, 0.2)
```
</div>
</div>
</div>

<div class = "exercises">
**Exercise E.2**   

What should be in empty spaces in order to write the likelihood function in R.

<div class="collapsibleSolution">
<button class="trigger">Solution</button>
<div class="content">

```{r}
lh_ttest <- function(y0, y2, mu, sigma, delta) {
  prod(dnorm(y0, mu, sigma, log = F)) * prod(dnorm(y1, mu + delta, sigma, log = F))
}
```
</div>
</div>
</div>

<div class = "exercises">  
**Exercise E.3**   

You want to check if a coin is fair. You toss it for 100 times and 22 times get heads. Now you can compute if likelyhood (p==0.5) is accurate for it. Try it out in your R console!

<div class="collapsibleSolution">
<button class="trigger">Solution</button>
<div class="content">

ll_coin <- dbinom(22, size=100, prob=0.5, log=FALSE)

</div>
</div>
</div>

<div class = "exercises">  
**Exercise E.4**   

You want to run a Bayesian model to calculate how many pizzas to order for a party. Which prior would You use if:

    a. only close friends are coming and you know their habits
    b. not that close friends are coming and you have a rough idea of their habits
    c. you invited new acquaintances
    
<div class="collapsibleSolution">
<button class="trigger">Solution</button>
<div class="content">


a. strongly informative


b. weakly informative


c. uninformative prior

</div>
</div>
</div>

<!-- Maybe for HW? -->
<!-- <div class = "exercises"> -->
<!-- **Exercise E.8.5**    -->
<!-- You want to approximate how many people visited your webpage for the first time today.  You know that overall 50 people were there. And you see 7 comments, five of which are left by your old followers. Print the most likely number of new visitors.  -->
<!-- <div class="collapsibleSolution"> -->
<!-- <button class="trigger">Solution</button> -->
<!-- <div class="content"> -->
<!-- ```{r} -->

<!-- # ll_new_visitors <- tibble( -->
<!-- #     m = 2:45, -->
<!-- #     LH = dhyper(2, m, 50,7) -->
<!-- # ) %>% -->
<!-- # which(ll_new_visitors==max(ll_new_visitors["LH"])) -->


<!-- </div> -->
<!-- </div> -->
<!-- </div> -->

<div class = "exercises">
**Exercise E.5**

TRUE or FAULSE?

a. For the Binomial model observed variables are number of trials and the probability of success.
b. In the graphical representation for the T-test model $g_i$ stands for $"group_i"$
c. Binomial Model can have only discrete observed variables. 
d. In a Linear Regression Model $y_i$ id determened by it's underlying distribution.

<div class="collapsibleSolution">
<button class="trigger">Solution</button>
<div class="content">
a. False
b. True
c. True
d. False

</div >
</div >
</div >

<div class = "exercises">

**Exercise E.6**   

Which model would you rather use do investigate the following:

a. If there  is a difference between chocolate consumption per person in Germany and Belgium?

b. You want to find out how many people visited your webpage for the first time.  You know that overall  50 people were there. And you see 7 comments, five of which are left by your old followers.  

c. If running speed increases with every 100m among sprinters, but not within marathoners.

d. Relationship between time spent on studying and the amount of coffee consumption.

e. If a football referee biased and gives yellow cards more often to one of the teems. Consider that the fold probability is known for each team.
 
<div class="collapsibleSolution">
<button class="trigger">Solution</button>
<div class="content">

a. T-Test Model
b. Flip-and-Draw-Hypergeometric Model
c. Linear Regression with Two Groups
d. Simple Linear Regression Model
e. Binomial
</div>
</div>
</div>



<div class = "exercises">
**Exercise E.7**

TRUE or FALSE? 

a. MLE (maximum likelihood estimate) is a method to check which model is more plausible


b. MLE (maximum likelihood estimate) can help in predicting data when the model is given


c. AIC (Akaike information criterion) can be used for model comparison


d. With Bayesian approach one cannot perform a parameter estimation


<div class="collapsibleSolution">
<button class="trigger">Solution</button>
<div class="content">

a. False

b. False

c. True

d. False

</div>
</div>
</div>


## Chapter 9

<div class = "exercises">


**Exercise E.1**  

Based on notation above, which of the following statements you think is true?

a. prior always influence the posteror more, than likelihood

b. the less the certainty in the prior, the more  posterior influenced by it 

c. the posterior is greater influenced by the likelihood if there is less certainty in the prior

d. likelihood always influence the posteror more, than the prior 


<div class="collapsibleSolution">
<button class="trigger">Solution</button>
<div class="content">


a. False

b. False

c. True

d. False

</div>
</div>
</div>


<div class = "exercises">
**Exercise E.2**  

How should you fill the code to get the plot of the posterior distribution for the coin flip scenario with k=6, N=24?

```{r, eval=F}
theta = seq(0, 1, length.out = 401)
likelihood <-dbinom(6, 24, theta) 
prior <-dbeta(theta, 1, 1)

posterior_plot <- as_tibble(theta) %>%
  mutate(posterior <- ____ * ____) %>%
  ggplot(aes(___, posterior))+
   geom_line()+
posterior_plot
```




<div class="collapsibleSolution">
<button class="trigger">Solution</button>
<div class="content">

```{r, eval=F}
theta = seq(0, 1, length.out = 401)
likelihood <- dbinom(6, 24, theta)
prior <- dbeta(theta, 1, 1)
posterior_plot <- as_tibble(theta) %>%
  mutate(posterior <- likelihood * prior) %>%
  ggplot(aes(theta, posterior))+
   geom_line()+
posterior_plot
```

</div>
</div>
</div>


<div class = "exercises">
**Exercise E.3**   

Can you think of a situation, when MLE and MAP output are exactly the same? HINT: Think which prior eliminates the difference between them!

<div class="collapsibleSolution">
<button class="trigger">Solution</button>
<div class="content">

MLE is a special case of MAP with an uniform the prior.

</div>
</div>
</div>

<div class = "exercises">
**Exercise E.4**   

Suppose Jones flipped a coin with unknown bias 30 times. She observed 20 heads. She updates her beliefs rationally with Bayes rule. Her posterior beliefs have the form of a beta distribution with parameters a = 25, b = 15. What distribution and what parameter values of that distribution capture Jones' prior beliefs before updating her beliefs with this data?

<div class="collapsibleSolution">
<button class="trigger">Solution</button>
<div class="content">

 $\text{Beta}(5,5)$

</div>
</div>
</div>




<div class = "exercises">
**Exercise E.5** 

Pick one correct answer:
The most frequently used point-estimate of Bayesian parameter estimation looks at 
a. the median of the a posterior distribution

b. the maximum likelihood estimate

c. the mean of the posterior distribution

d. the normalizing constant in Bayes rule 

<div class="collapsibleSolution">
<button class="trigger">Solution</button>
<div class="content">

b.

</div>
</div>
</div>


<div class = "exercises">

**Exercise E.6** 

The most frequently used interval-based estimate in frequentist approaches is:

a. the support of the likelihood distribution

b. the credible interval

c. the hypothesis interval

d. the 95% highest-density interval of the maximum likelihood estimate

<div class="collapsibleSolution">
<button class="trigger">Solution</button>
<div class="content">

b.

</div>
</div>
</div>

<div class = "exercises">
**Exercise E.7**

In which order should the listed Markov Chain Monte Carlo steps be?

d. Set an initial value

g. Generate a new value (proposal)

a. Compare the posterior value of the new proposal and the height of the posterior at the previous step

c.If the new proposal has a higher posterior value than the most recent sample, then accept the new proposal.

f. if the new proposal has a lower posterior value than the most recent sample, then b.. compute the proportion of the posterior value of the new proposal and the height of the posterior at the previous step

e. choose to accept or reject the new proposal in respect to the computed proportion


<div class="collapsibleSolution">
<button class="trigger">Solution</button>
<div class="content">

d,g,a,c,f,b

</div >






